### 2013-05-23 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding more rda analysis

%%---------------------------------------------------------------------------%%
\section{Advanced Preconditioning Strategies}
\label{subsec:spn_advanced_preconditioning}
For the fuel assembly criticality problem presented in the previous
section, the spectral radius was near one and in a regime in which
MCSA breakdown is observed. In this regime the number of stochastic
histories required to converge MCSA increases rapidly and the
resulting poor performance is compounded by those histories being
increasingly expensive to compute. To overcome this difficulty,
advanced preconditioning strategies for the $SP_N$ problem are
required beyond simple Jacobi methods that can reduce the spectral
radius into a region of better MCSA performance. Several modern
algebraic preconditioning strategies will be presented here and used
within the explicit preconditioning framework given by
Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on MCSA
solutions of the fuel assembly criticality problem will be presented.

@@

MCSA preconditioned with ILUT was used to solve the fuel assembly
problem presented in the previous section for a 1-group $SP_1$
discretization. Unlike the Jacobi preconditioned strategy, convergence
was achieved with ILUT preconditioning. To study convergence
sensitivity to ILUT parameters, the fill level and drop tolerance were
parametrically varied with the number of iterations to converge the
second eigenvalue iteration for the fuel assembly problem recorded
along with the maximum number of non-zero entries observed in all
matrix rows for the left/right preconditioned composite linear
operator. To provide some sparsity to the factorization, the ILUT drop
tolerance was used to drop elements in the extracted inverse
triangular factors. For each calculation, the number of iterations
required to converge reported was for the first eigenvalue iteration
with \sn{3}{4} histories at each MCSA iteration to compute the Monte
Carlo correction using the adjoint collision estimator. All ILUT
calculations reported here were performed on a single CPU and
therefore this data does not take into account the aforementioned
effects of parallel decomposition on the quality of the ILUT
preconditioning.

@@

Unfortunately, gaining convergence (and excellent iterative
performance) with MCSA for the $SP_N$ fuel assembly problem comes at
an immediate cost. Figure~\ref{fig:ilut_size} gives the maximum number
of non-zero entries in the composite linear operator generated by
preconditioning as a function ILUT drop tolerance for varying values
of fill level. For the 1-group $SP_1$ discretization, the original
linear operator will contain only a maximum of 7 non-zero entries per
row in the system. As observed in Figure~\ref{fig:ilut_size},
performing the explicit preconditioning yields composite linear
operators with $O(1,000)$ elements in a row for all combinations of
fill level and drop tolerance, over 10\% of the row size for this
particular. This massive number of row entries, observed for a
significant fraction of rows in the system, creates several
problems. First, sparsity is completely destroyed with each state in
the system now coupled to over 10\% of the total states in the system
through the composite iteration matrix. This means that Monte Carlo
sampling tables will be large, requiring significant memory to store
them and a substantial overhead in the sampling procedure during
transport. In addition, the matrix-matrix multiply operations required
to build the composite operator see significant performance losses due
to the amount of data that must be handled. In parallel, losing
sparsity severely inhibits performance where now parallel operations
require communications amongst orders of magnitude more processors for
nearest-neighbor type algorithms.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/ilut_size.pdf}
  \end{center}
  \caption{\textbf{Maximum number of non-zero entries observed for all
      rows in the composite linear operator for the fuel assembly
      problem with ILUT preconditioning given as a function of ILUT
      drop tolerance.}  \textit{Each colored curve represents the row
      size for a different ILUT fill level. Fill levels of 1.5, 2.0,
      2.5, and 3.0 were used.}}
  \label{fig:ilut_size}
\end{figure}

As a means to assess the quality of the preconditioning, a simple
metric is developed to address these concerns and allow comparison to
future developments. For our studies, our core performance metric will
be iterative performance with the minimum number of MCSA iterations
required to converge the problem desired. For Monte Carlo
calculations, this improved performance is balanced by the creation of
a dense composite system and we seek to reduce the number of non-zero
entries to a minimum value in order to lower the amount of coupling
amongst states in the system and reduce the amount of memory used
along with the potentially latency overhead. Given these two
objectives, the following metric is proposed:
\begin{equation}
  \text{Quality Metric} = (\text{\# iterations})*(\text{maximum \# of
    non-zero values})\:,
\end{equation}
where the highest quality preconditioning is one that minimizes this
metric. For the ILUT preconditioned data provided,
Figure~\ref{fig:ilut_quality} provides the computed metric as a
function of ILUT drop tolerance for each fill level used. In general,
the metric is similar to the data observed for the number of
iterations required to converge as the non-zero row entries changed by
a smaller fraction over drop tolerances tested.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/ilut_quality.pdf}
  \end{center}
  \caption{\textbf{ILUT preconditioning quality metric for the fuel
      assembly problem given as a function of ILUT drop tolerance.}
    \textit{Each colored curve represents the quality metric behavior
      for a different ILUT fill level. Fill levels of 1.5, 2.0, 2.5,
      and 3.0 were used.}}
  \label{fig:ilut_quality}
\end{figure}

@@

\subsection{Sparse Approximate Inverse Preconditioning}
\label{subsec:spn_spainv_preconditioning}
Explicit formation of the preconditioner inverse is a requirement of
the chosen MCSA preconditioning strategy. As a result, although ILUT
preconditioning provided excellent iterative performance, a
significant time penalty is paid to extract the inverse of the
triangular factors through $N$ triangular solves. In addition, this
explicit inversion yielded a composite linear operator that was dense,
requiring modification of the system in order to achieve better
scalability. Sparse approximate inverse (SPAINV) preconditioning is a
technique that may potentially alleviate both of these constraints at
the cost of reduced iterative performance. The method produces a
sparse approximation of the inverse matrix directly by minimizing the
Frobenius norm of the residual matrix \citep{saad_iterative_2003}:
\begin{equation}
  || \mathbf{I} - \mathbf{A} \mathbf{M} ||^2_F =
  \sum_{j=1}^N ||\mathbf{e}_j - \mathbf{A} \mathbf{m}_j||^2_2 \:,
  \label{eq:frobenius_norm_min}
\end{equation}
where the inverse preconditioning matrix $\mathbf{M}$ minimizes the
norm on the right and $\mathbf{e}_j$ and $\mathbf{m}_j$ are the
$j^{th}$ columns of the identity and preconditioning matrices
respectively. Here, $\mathbf{M}$ is the actual inverse of the
preconditioner and is formed directly by the algorithm. On the right
hand side of Eq~(\ref{eq:frobenius_norm_min}), the Frobenius norm is
represented as a effective sum of $N$ linear system residual norms. A
few iterations of a subspace method or other iterative scheme can be
used to effectively solve these systems to a loose tolerance and yield
the columns of the approximate preconditioning matrix. At each
iteration, threshold values are used to eliminate the components of
each $\mathbf{m}_j$ column to maintain the desired level of
sparsity. In addition, a sparsity pattern can be predetermined and
enforced during this process. For the SPAINV implementation used for
this work, the sparsity pattern was defined by levels such that a
level of $N$ for an input operator $\mathbf{A}$ will generate a
preconditioning matrix $\mathbf{M}$ with the same sparsity pattern as
$\mathbf{A}^{N+1}$. By using the norm minimization strategy instead of
a factorization such as ILUT, parallel results are reproduced
regardless of parallel problem size, meaning that a serial computation
will converge in the same number of iterations as a computation with
thousands of cores.

MCSA preconditioned with SPAINV was used to solve the fuel assembly
problem as with ILUT for a 1-group $SP_1$ discretization. To study
convergence sensitivity to SPAINV parameters, the number of levels in
the sparsity pattern and threshold were parametrically varied with the
number of iterations to converge the second eigenvalue iteration for
the fuel assembly problem recorded along with the maximum number of
non-zero entries observed in all matrix rows for the left/right
preconditioned composite linear operator. For each calculation, the
number of iterations required to converge reported was for the first
eigenvalue iteration with \sn{3}{4} histories at each MCSA iteration
to compute the Monte Carlo correction using the adjoint collision
estimator. All SPAINV calculations reported here were performed on a
single CPU.

@@

\subsection{SPAINV Improved ILUT Preconditioning}
\label{subsec:spn_spainv_preconditioning}
For improved convergence over SPAINV preconditioning while maintaining
sparsity, ILUT preconditioning may be improved through an additional
application of SPAINV (SPAINV-ILUT)\citep{saad_iterative_2003}. For
the ILUT preconditioning case we have the following linear problem to
solve for the substituted variable:
\begin{equation}
  \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
  \mathbf{L}^{-1} \mathbf{b}\:.
  \label{eq:ilut_standalone}
\end{equation}
We then apply SPAINV and minimize the following problem, this time on
the left:
\begin{equation}
  || \mathbf{I} - \mathbf{M} \mathbf{L}^{-1} \mathbf{A}
  \mathbf{U}^{-1} ||^2_F\:,
  \label{eq:ilut_minimize}
\end{equation}
giving a new preconditioned system:
\begin{equation}
  \mathbf{M} \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
  \mathbf{M} \mathbf{L}^{-1} \mathbf{b}\:.
  \label{eq:ilut_spainv}
\end{equation}
Doing this permits a smaller fill level to be used with ILUT and a
lower-order sparsity pattern to be used with SPAINV, giving more
sparsity in the resulting composite linear operator while maintaining
good iterative performance.

MCSA preconditioned with SPAINV-ILUT was also used to solve the same
fuel assembly problem as with the other preconditioners for the same
1-group $SP_1$ discretization. To study convergence sensitivity to
SPAINV parameters, the ILUT drop tolerance and ILUT fill level were
parametrically varied with the number of iterations to converge the
second eigenvalue iteration for the fuel assembly problem recorded
along with the maximum number of non-zero entries observed in all
matrix rows for the left/right preconditioned composite linear
operator. For each calculation, the number of iterations required to
converge reported was for the first eigenvalue iteration with
\sn{3}{4} histories at each MCSA iteration to compute the Monte Carlo
correction using the adjoint collision estimator. All SPAINV-ILUT
calculations reported here were performed on a single
CPU.

Figure~\ref{fig:spainv_ilut_iterations} gives the number of iterations
required to converge as a function of ILUT drop tolerance for ILUT
fill levels of 1.5, 2.0, 3.0, 4.0 and 5.0 with a fixed SPAINV
threshold of 1.0 and sparsity level of 1. Threshold values and higher
resorted in loss of convergence while larger sparsity levels resulted
in composite operator that was too dense. At smaller levels of fill,
the benefits of using an addition step of SPAINV preconditioning are
readily observed with the number of iterations required to converge
cut in half for large drop tolerances. For smaller drop tolerances and
higher ILUT levels of fill, the additional step of SPAINV
preconditioning offers marginal improvement to iterative
performance. As shown in Figure~\ref{fig:spainv_ilut_size},
because ILUT is used in the preconditioning sequence, sparsity is
again loss with only marginally improved non-zero entry numbers due to
the application of the SPAINV threshold. As a result, the quality
metrics given by Figure~\ref{fig:spainv_ilut_quality} are only
marginally better than those observed for ILUT preconditioning and
much larger than the SPAINV quality metric values.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/psilut_iterations.pdf}
  \end{center}
  \caption{\textbf{Number of MCSA iterations required to converge the
      second eigenvalue iteration for the fuel assembly problem with
      SPAINV-ILUT preconditioning as a function of ILUT drop
      tolerance.} \textit{Each colored curve represents the row size
      for a different ILUT fill level. Fill levels of 1.5, 2.0, 3.0,
      4.0, and 5.0 were used.}}
  \label{fig:spainv_ilut_iterations}
\end{figure}

@@

\subsection{Applying the Reduced Domain Approximation}
\label{subsec:spn_prec_rda}
For each preconditioning technique presented, although convergence was
achieved a primary concern is the number of non-zero states in each
row of the system generated by the explicit preconditioning
strategy. In many cases, orders of magnitude more matrix elements were
generated resulting in poor scalability for domain decomposed
algorithms. As outlined in
\S~\ref{subsec:reduced_domain_approximation}, the reduced domain
approximation may be used as a mechanism to potentially alleviate this
problem by filtering elements of the composite matrix in each row that
fall below a certain threshold value or by maintaining the largest $N$
elements in each row where $N$ is a designated fill level.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/rda_iterations.pdf}
  \end{center}
  \caption{\textbf{Number of MCSA iterations required to converge the
      second eigenvalue iteration for the fuel assembly problem with
      preconditioning as a function of reduced domain approximation
      fill level.}  \textit{The largest fill level for each
      preconditioning presented is that using the parameters that gave
      the best results without the approximation.}}
  \label{fig:rda_iterations}
\end{figure}

@@

\subsection{MCSA Relaxation Parameters}
\label{subsec:spn_mcsa_relaxation}
The Richardson iteration upon which the Neumann-Ulam method is built
may be implemented with a scalar relaxation parameter that can be
adjusted to improve convergence:
\begin{equation}
  \mathbf{x} = \mathbf{x} + \omega \mathbf{r}\:,
  \label{eq:richardson_relaxation}
\end{equation}
where $\omega$ is the relaxation parameter. Analogously, these
relaxation parameter techniques can be applied to Monte Carlo to
improve convergence as demonstrated by Dimov
\citep{dimov_new_1998}. In this case, building an iteration matrix
from Eq~(\ref{eq:richardson_relaxation} gives:
\begin{equation}
  \mathbf{H} = \mathbf{I} - \omega \mathbf{A}\:,
  \label{eq:relaxed_iteration_matrix}
\end{equation}
with the probabilities and weights for the Monte Carlo game
appropriately scaled. By inspection, such a scaling is a simple form
of preconditioning on the left where all rows in the system are scaled
by the same scalar parameter. For MCSA, we can stage this scheme with
two separate relaxation parameters: one for the outer Richardson
iteration and one for the inner Monte Carlo solve:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \omega_R \ve{r}^k\:,\\ \ve{r}^{k+1/2}
    = \ve{b} -
    \ve{A}\ve{x}^{k+1/2}\:,\\ \omega_N\ve{A}\delta\ve{x}^{k+1/2} =
    \omega_N\ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
    \ve{x}^{k+1/2}\:,\\ \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:mcsa_relaxed}
\end{subequations}
where $\omega_R$ is the Richardson iteration relaxation parameter and
$\omega_N$ is the Neumann-Ulam Monte Carlo solve relaxation parameter.

>>>>

%%---------------------------------------------------------------------------%%
\section{Advanced Preconditioning Strategies}
\label{subsec:spn_advanced_preconditioning}
For the fuel assembly criticality problem presented in the previous
section, the spectral radius was near one using Jacobi preconditioning
and in a regime in which MCSA breakdown is observed. In this regime
the number of stochastic histories required to converge MCSA increases
rapidly and the resulting poor performance is compounded by those
histories being increasingly expensive to compute. To overcome this
difficulty, advanced preconditioning strategies for the $SP_N$ problem
are required beyond simple Jacobi methods that can reduce the spectral
radius into a region of better MCSA performance. Several modern
algebraic preconditioning strategies will be presented here and used
within the explicit preconditioning framework given by
Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on MCSA
solutions of the fuel assembly criticality problem will be presented.

@@

MCSA preconditioned with ILUT was used to solve the fuel assembly
problem presented in the previous section for a 1-group $SP_1$
discretization. Unlike the Jacobi preconditioned strategy, convergence
was achieved with ILUT preconditioning. To study convergence
sensitivity to ILUT parameters, the fill level and drop tolerance were
parametrically varied with the number of iterations to converge the
second eigenvalue iteration for the fuel assembly problem recorded
along with the maximum number of non-zero entries observed in all
matrix rows for the left/right preconditioned composite linear
operator. To provide some sparsity to the factorization, the ILUT drop
tolerance was used to drop elements in the extracted inverse
triangular factors. For each calculation, the number of iterations
required to converge reported was for the secon deigenvalue iteration
with \sn{3}{4} histories at each MCSA iteration to compute the Monte
Carlo correction using the adjoint collision estimator. All ILUT
calculations reported here were performed on a single CPU and
therefore this data does not take into account the aforementioned
effects of parallel decomposition on the quality of the ILUT
preconditioning.

@@

Unfortunately, gaining convergence (and excellent iterative
performance) with MCSA for the $SP_N$ fuel assembly problem comes at
an immediate cost. Figure~\ref{fig:ilut_size} gives the maximum number
of non-zero entries in the composite linear operator generated by
preconditioning as a function ILUT drop tolerance for varying values
of fill level. For the 1-group $SP_1$ discretization, the original
linear operator will contain only a maximum of 7 non-zero entries per
row in the system. As observed in Figure~\ref{fig:ilut_size},
performing the explicit preconditioning yields composite linear
operators with $O(1,000)$ elements in a row for all combinations of
fill level and drop tolerance, over 10\% of the total row size for
this particular problem. This large number of row entries, observed
for a significant fraction of rows in the system, creates several
problems. First, sparsity is completely destroyed with each state in
the system now coupled to over 10\% of the total states in the system
through the composite iteration matrix. This means that Monte Carlo
sampling tables will be large, requiring significant memory to store
them and a substantial overhead in the sampling procedure during
transport. In addition, the matrix-matrix multiply operations required
to build the composite operator see significant performance losses due
to the amount of data that must be handled. In parallel, losing
sparsity severely inhibits performance where now parallel operations
require communications amongst orders of magnitude more processors for
nearest-neighbor type algorithms.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/ilut_size.pdf}
  \end{center}
  \caption{\textbf{Maximum number of non-zero entries observed for all
      rows in the composite linear operator for the fuel assembly
      problem with ILUT preconditioning given as a function of ILUT
      drop tolerance.}  \textit{Each colored curve represents the row
      size for a different ILUT fill level. Fill levels of 1.5, 2.0,
      2.5, and 3.0 were used.}}
  \label{fig:ilut_size}
\end{figure}

As a means to assess the quality of the preconditioning, a simple
metric is developed to address these concerns and allow comparison to
future developments. For our studies, our core performance metric will
be iterative performance with the minimum number of MCSA iterations
required to converge the problem desired. For Monte Carlo
calculations, this improved performance is balanced by the creation of
a dense composite system and we seek to reduce the number of non-zero
entries to a minimum value in order to lower the amount of coupling
amongst states in the system and reduce the amount of memory used
along with the potentially latency overhead. Given these two
objectives, the following metric is proposed:
\begin{equation}
  \text{Quality Metric} = (\text{\# iterations}) \times (\text{maximum
    \# of non-zero values})\:,
\end{equation}
where the highest quality preconditioning is one that minimizes this
metric. For the ILUT preconditioned data provided,
Figure~\ref{fig:ilut_quality} provides the computed metric as a
function of ILUT drop tolerance for each fill level used. In general,
the metric is similar to the data observed for the number of
iterations required to converge as the non-zero row entries changed by
a smaller fraction over drop tolerances tested.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/ilut_quality.pdf}
  \end{center}
  \caption{\textbf{ILUT preconditioning quality metric for the fuel
      assembly problem given as a function of ILUT drop tolerance.}
    \textit{Each colored curve represents the quality metric behavior
      for a different ILUT fill level. Fill levels of 1.5, 2.0, 2.5,
      and 3.0 were used.}}
  \label{fig:ilut_quality}
\end{figure}

@@

\subsection{Sparse Approximate Inverse Preconditioning}
\label{subsec:spn_spainv_preconditioning}
Explicit formation of the preconditioner inverse is a requirement of
the chosen MCSA preconditioning strategy. As a result, although ILUT
preconditioning provided excellent iterative performance, a
significant time penalty is paid to extract the inverse of the
triangular factors through $N$ triangular solves. In addition, this
explicit inversion yielded a composite linear operator that was dense,
requiring modification of the system in order to achieve better
scalability and Monte Carlo performance. Sparse approximate inverse
(SPAINV) preconditioning is a technique that may potentially alleviate
both of these constraints at the cost of reduced iterative
performance. The method produces a sparse approximation of the inverse
matrix directly by minimizing the Frobenius norm of the residual
matrix \citep{saad_iterative_2003}:
\begin{equation}
  || \mathbf{I} - \mathbf{A} \mathbf{M} ||^2_F =
  \sum_{j=1}^N ||\mathbf{e}_j - \mathbf{A} \mathbf{m}_j||^2_2 \:,
  \label{eq:frobenius_norm_min}
\end{equation}
where the inverse preconditioning matrix $\mathbf{M}$ minimizes the
norm on the right and $\mathbf{e}_j$ and $\mathbf{m}_j$ are the
$j^{th}$ columns of the identity and preconditioning matrices
respectively. Here, $\mathbf{M}$ is the actual inverse of the
preconditioner and is formed directly by the algorithm. On the right
hand side of Eq~(\ref{eq:frobenius_norm_min}), the Frobenius norm is
represented as a effective sum of $N$ linear system residual norms. A
few iterations of a subspace method or other iterative scheme can be
used to effectively solve these systems to a loose tolerance and yield
the columns of the approximate preconditioning matrix. At each
iteration, threshold values are used to eliminate the components of
each $\mathbf{m}_j$ column to maintain the desired level of
sparsity. In addition, a sparsity pattern can be predetermined and
enforced during this process. For the SPAINV implementation used for
this work, the sparsity pattern was defined by levels such that a
level of $N$ for an input operator $\mathbf{A}$ will generate a
preconditioning matrix $\mathbf{M}$ with the same sparsity pattern as
$\mathbf{A}^{N+1}$. By using the norm minimization strategy instead of
a factorization such as ILUT, parallel results are reproduced
regardless of parallel problem size, meaning that a serial computation
will converge in the same number of iterations as a computation with
thousands of cores.

MCSA preconditioned with SPAINV was used to solve the fuel assembly
problem as with ILUT for a 1-group $SP_1$ discretization. To study
convergence sensitivity to SPAINV parameters, the number of levels in
the sparsity pattern and threshold were parametrically varied with the
number of iterations to converge the second eigenvalue iteration for
the fuel assembly problem recorded along with the maximum number of
non-zero entries observed in all matrix rows for the left/right
preconditioned composite linear operator. For each calculation, the
number of iterations required to converge reported was for the second
eigenvalue iteration also with \sn{3}{4} histories at each MCSA
iteration to compute the Monte Carlo correction using the adjoint
collision estimator. All SPAINV calculations reported here were
performed on a single CPU.

@@

\subsection{SPAINV Improved ILUT Preconditioning}
\label{subsec:spn_spainv_preconditioning}
As a means to improve convergence over using only SPAINV
preconditioning, ILUT preconditioning may be improved through an
additional application of SPAINV
(SPAINV-ILUT)\citep{saad_iterative_2003}. For the ILUT preconditioning
case we have the following linear problem to solve for the substituted
variable:
\begin{equation}
  \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
  \mathbf{L}^{-1} \mathbf{b}\:.
  \label{eq:ilut_standalone}
\end{equation}
We then apply SPAINV and minimize the following problem, this time on
the left:
\begin{equation}
  || \mathbf{I} - \mathbf{M} \mathbf{L}^{-1} \mathbf{A}
  \mathbf{U}^{-1} ||^2_F\:,
  \label{eq:ilut_minimize}
\end{equation}
giving a new preconditioned system:
\begin{equation}
  \mathbf{M} \mathbf{L}^{-1} \mathbf{A} \mathbf{U}^{-1} \mathbf{u} =
  \mathbf{M} \mathbf{L}^{-1} \mathbf{b}\:.
  \label{eq:ilut_spainv}
\end{equation}
Doing this permits a smaller fill level to be used with ILUT and a
lower-order sparsity pattern to be used with SPAINV, giving marginally
more sparsity in the resulting composite linear operator while
maintaining good iterative performance.

MCSA preconditioned with SPAINV-ILUT was also used to solve the same
fuel assembly problem as with the other preconditioners for the same
1-group $SP_1$ discretization. To study convergence sensitivity to
SPAINV-ILUT parameters, the ILUT drop tolerance and ILUT fill level
were parametrically varied with the number of iterations to converge
the second eigenvalue iteration for the fuel assembly problem recorded
along with the maximum number of non-zero entries observed in all
matrix rows for the left/right preconditioned composite linear
operator. SPAINV parameters for the data presented here were fixed as
it was found that the preconditioning quality was largely insenstive
to their variation. In addition, the fact that SPAINV is being used to
preconditioned a problem already preconditioned with ILUT means that
the input matrix sparsity pattern is already somewhat dense and large
SPAINV levels would result in an even denser system. Therefore, a
smaller sparsity pattern level will be used for this
preconditioning. For each calculation, the number of iterations
required to converge reported was for the second eigenvalue iteration.
Again, \sn{3}{4} histories are used at each MCSA iteration to compute
the Monte Carlo correction using the adjoint collision estimator. All
SPAINV-ILUT calculations reported here were performed on a single CPU.

Figure~\ref{fig:spainv_ilut_iterations} gives the number of iterations
required to converge as a function of ILUT drop tolerance for ILUT
fill levels of 1.5, 2.0, 3.0, 4.0 and 5.0 with a fixed SPAINV
threshold of 1.0 and sparsity level of 1. Threshold values higher than
1.0 resulted in a loss of convergence and lower values did not
significantly affect performance. Larger sparsity levels resulted in a
composite operator that was too dense and a prohibitive amount of CPU
time required to compute the preconditioner. At smaller levels of
fill, the benefits of using an additional step of SPAINV
preconditioning are readily observed with the number of iterations
required to converge cut in half for large drop tolerances. For
smaller drop tolerances and higher ILUT levels of fill, the additional
step of SPAINV preconditioning offers marginal improvement to
iterative performance. As shown in Figure~\ref{fig:spainv_ilut_size},
because ILUT is used in the preconditioning sequence, sparsity is
again lost with only marginally improved non-zero entry numbers due to
the application of the SPAINV threshold. As a result, the quality
metrics given by Figure~\ref{fig:spainv_ilut_quality} are only
marginally better than those observed for ILUT preconditioning and
much larger than the SPAINV quality metric values.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/psilut_iterations.pdf}
  \end{center}
  \caption{\textbf{Number of MCSA iterations required to converge the
      second eigenvalue iteration for the fuel assembly problem with
      SPAINV-ILUT preconditioning as a function of ILUT drop
      tolerance.} \textit{Each colored curve represents the row size
      for a different ILUT fill level. Fill levels of 1.5, 2.0, 3.0,
      4.0, and 5.0 were used.}}
  \label{fig:spainv_ilut_iterations}
\end{figure}

@@

\subsection{Applying the Reduced Domain Approximation}
\label{subsec:spn_prec_rda}
For each preconditioning technique presented convergence was achieved
for the single fuel assembly problem. However, a primary concern is
the number of non-zero states in each row of the system generated by
the explicit preconditioning strategy. In many cases, orders of
magnitude more matrix elements were generated resulting in poor
scalability for domain decomposed algorithms and overall performance
issues for Monte Carlo. As outlined in
\S~\ref{subsec:reduced_domain_approximation}, the reduced domain
approximation may be used as a mechanism to potentially alleviate this
problem by filtering elements of the composite matrix in each row that
fall below a certain threshold value or by maintaining the largest $N$
elements in each row where $N$ is a designated fill level.

For the ILUT, SPAINV, and SPAINV-ILUT preconditioning strategies the
reduced domain approximation will be applied to reduce the density of
the composite linear operator to more manageable levels. For each
preconditioner, the parameters that achieved the best quality metric
results from the previous analysis were used with the reduced domain
threshold was set to \sn{1}{-10} in order to eliminate any exceedingly
small values from the matrix (generally this is simply removing
non-zero elements within the floating point tolerance of zero). The
reduced domain fill level was then varied, starting with the largest
non-zero entries per row value observed for each of the
preconditioning types in order to assess its effects relative to the
case where no reduced domain approximation was applied.

Figure~\ref{fig:rda_iterations} gives the number of iterations
required to converge for each preconditioner type as a function of
reduced domain fill level. Figure~\ref{fig:rda_quality} gives the
corresponding quality metric for each data point where the number of
non-zero entries used to compute the metric is equivalent to the
reduced domain fill level. We first note the SPAINV preconditioning
alone is signficantly more sensitive to the reduction in domain size
over the ILUT-based methods, although the preconditioner was of
$O(100)$ non-zero entries per row without any approximation
applied. For the ILUT-based methods, performance was significantly
better with converge achieved in less than 40 MCSA iterations with
only 10 non-zero entries in each row (vs. 7 non-zero entries for the
case with no preconditioning). SPAINV-ILUT iterative performance was
marginally better than ILUT alone for all reduced domain fill
levels. However, it should be noted that the ILUT level of fill used
for the ILUT only calculations was set to 4 for this case while the
SPAINV-ILUT preconditioner used an ILUT level of fill of 5 and
therefore the marginally better performance is more likely a result of
this addition of fill rather than the extra SPAINV
preconditioning. Looking at the quality metric data, we see a nice
power-law reduction in the quality metric as a function of the reduced
domain fill level, achieving 2 orders of magnitude reduction in the
quality metric for the ILUT-based preconditioning methods.

Although applying the reduced domain approximation results in a
successful recovery of sparsity for the Monte Carlo problem while
maintaining good convergence properties, there is still an issue of
forming the composite operator before applying the approximation and
potentially generating the transpose of this operator in the case of
the adjoint Monte Carlo method. Because of this, memory and scaling
issues may still be observed when building the Monte Carlo problem. In
addition, the expensive extraction of the inverse of the
preconditioning operators for the explicit scheme creates a
significant cost in overall performance. Future work in this area
should be considerate of these important components of the algorithm.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/rda_iterations.pdf}
  \end{center}
  \caption{\textbf{Number of MCSA iterations required to converge the
      second eigenvalue iteration for the fuel assembly problem with
      each preconditioning as a function of reduced domain
      approximation fill level.}  \textit{The largest fill level for
      each preconditioning presented is that using the parameters that
      gave the best results without the approximation.}}
  \label{fig:rda_iterations}
\end{figure}

@@

\subsection{MCSA Relaxation Parameters}
\label{subsec:spn_mcsa_relaxation}
The Richardson iteration upon which the Neumann-Ulam method is built
may be implemented with a scalar relaxation parameter that can be
adjusted to improve convergence:
\begin{equation}
  \mathbf{x} = \mathbf{x} + \omega \mathbf{r}\:,
  \label{eq:richardson_relaxation}
\end{equation}
where $\omega$ is the relaxation parameter. This is effectively a form
of preconditioning where the system is being scaled on the left by a
constant value in all rows. Analogously, these relaxation parameter
techniques can be applied to Monte Carlo to improve convergence as
demonstrated by Dimov \citep{dimov_new_1998}. In this case, building
an iteration matrix from Eq~(\ref{eq:richardson_relaxation}) gives:
\begin{equation}
  \mathbf{H} = \mathbf{I} - \omega \mathbf{A}\:,
  \label{eq:relaxed_iteration_matrix}
\end{equation}
with the probabilities and weights for the Monte Carlo game
appropriately scaled. By inspection, such a scaling is a simple form
of preconditioning on the left where all rows in the system are scaled
by the same scalar parameter. For MCSA, we can stage this scheme with
two separate relaxation parameters; one for the outer Richardson
iteration and one for the inner Monte Carlo solve:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \omega_R \ve{r}^k\:,\\ \ve{r}^{k+1/2}
    = \ve{b} -
    \ve{A}\ve{x}^{k+1/2}\:,\\ \omega_N\ve{A}\delta\ve{x}^{k+1/2} =
    \omega_N\ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
    \ve{x}^{k+1/2}\:,\\ \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:mcsa_relaxed}
\end{subequations}
where $\omega_R$ is the Richardson iteration relaxation parameter and
$\omega_N$ is the Neumann-Ulam Monte Carlo solve relaxation parameter.
